---
title: "Convert DMv4 to DMv6 coordinates"
author: "Kirk Amundson"
date: 2020_1030
output: html_notebook
---

Aim: Convert coordinates of the SolCAP 8303 SNP array to the coordinates of the DM1-3 v6.1 assembly

## Libraries:
```{r}
library(tidyverse)
library(readxl)
```

## Read in data:
### Flanking sequence of SolCAP 8303 array SNPs:
```{r}
solcap <- read_excel("../ref/12864_2011_3434_MOESM2_ESM.xlsx", skip = 2, col_names = c("solcap_id", "flanking_seq"))
head(solcap)
```

### DM1-3 v4 asembly SolCAP 8303 coordinates:
```{r}
v4_coords <- read_tsv("../ref/potato_8303SNPs_potato_dm_v4.03.gff3", 
                      col_names = c("chrom", "source", "type", "start", "end", "score", "strand", "phase", "attributes")) %>% 
  mutate(solcap_id = str_extract(attributes, "solcap_snp_c.+") %>% str_remove(";$")) %>% 
  group_by(solcap_id) %>%
  mutate(n = n()) %>% 
  filter(n == 1)
head(v4_coords)
```

> Problem: There are records in the GFF3 file that have the same name. Example: solcap_snp_c2_9926.
I'm going to throw these away before I do the BLAT search.

## Join flanking sequences to v4 coordinates, then convert to multi-FASTA for BLAST to v6
```{r}
add_seq <- v4_coords %>% 
  left_join(solcap) %>% 
  mutate(snp_index_seq = str_locate(flanking_seq, "\\[")) %>% 
  mutate(probe_for_blast = str_remove(flanking_seq, "\\[[ATCG]/") %>% str_remove("\\]")) %>% 
  mutate(blast_query = paste0(">", solcap_id, "_", chrom, "_", start, "_", end, "_flanking_seq\n", probe_for_blast))
```

## As a sanity check, blast these probe sequences to DM1-3 v4 and see if the recovered coordinates match up:
```{r}
write(add_seq$blast_query, file = "solcap_8303_context_sequences.fa")
```

```{r}
system("/Users/Kirk/miniconda3/bin/pblat ../ref/PGSC_DM_v4.03_pseudomolecules.fasta solcap_8303_context_sequences.fa solcap_8303_context_blat_to_DMv4.tsv -out=blast8")
```

```{r}
blast_cols <- c("query_id", "subject_id", "pct_identity", "aln_length",
                "n_mismatches", "n_gap_opens", "q_start", "q_end",
                "s_start", "s_end", "e_value", "bit_score")

v4_hits <- read_tsv("solcap_8303_context_blat_to_DMv4.tsv", col_names = blast_cols) %>% 
  arrange(subject_id, s_start) %>% 
  mutate(orig_chrom = str_extract(query_id, "[Cc]hr[0-9]{2}")) %>% 
  mutate(orig_start = str_extract(query_id, "[0-9]{1,}_flanking") %>% str_remove_all("_flanking")) %>% 
  mutate(s_chrom = str_extract(subject_id, "ch[0-9]{2}") %>% str_replace("ch", "chr")) %>% 
  mutate(s_left = pmin(s_start, s_end)) %>% 
  mutate(s_right = pmax(s_start, s_end)) %>% 
  mutate(pos_consistent = orig_start >= s_left & orig_start <= s_right) %>%
  mutate(chrom_consistent = orig_chrom == s_chrom)
```

```{r}
v4_hits_filtered <- v4_hits %>% 
  filter(aln_length >= 101) %>% 
  filter(n_mismatches <= 1) %>%
  filter(n_gap_opens == 0) %>%
  group_by(query_id) %>%
  mutate(n = n())
table(v4_hits_filtered$n)
```

> Sanity check filters kept 5,765 SNPs with single best hits in the 4.03 assembly. Apply these to DM1-3 v6. Filters are:
  1. Alignment length ≥ 90
  2. Only a single mismatch allowed
  3. No gaps allowed
  
```{r}
head(v4_hits_filtered)
v4_hits_filtered %>% 
  mutate(gff_from_left = s_left + 50,
         gff_from_right = s_right - 50,
         match_orig_left = gff_from_left == orig_start,
         match_orig_right = gff_from_right == orig_start) %>%
  group_by(match_orig_left, match_orig_right) 
```
  
## Lift over coordinates to DM1-3 v6
```{r}
system("/Users/Kirk/miniconda3/bin/pblat ../ref/DM_1-3_516_R44_potato_genome_assembly.v6.1.fa solcap_8303_context_sequences.fa solcap_8303_context_blat_to_DMv6.tsv -out=blast8")
```

```{r}
v6_hits <- read_tsv("solcap_8303_context_blat_to_DMv6.tsv", col_names = blast_cols) %>% 
  arrange(subject_id, s_start) %>% 
  mutate(orig_chrom = str_extract(query_id, "[Cc]hr[0-9]{2}")) %>% 
  mutate(orig_start = str_extract(query_id, "[0-9]{1,}_flanking") %>% str_remove_all("_flanking")) %>% 
  mutate(s_chrom = str_extract(subject_id, "ch[0-9]{2}") %>% str_replace("ch", "chr")) %>% 
  mutate(s_left = pmin(s_start, s_end)) %>% 
  mutate(s_right = pmax(s_start, s_end)) %>% 
  mutate(pos_consistent = orig_start >= s_left & orig_start <= s_right) %>%
  mutate(chrom_consistent = orig_chrom == subject_id)
```

```{r}
v6_hits_filtered <- v6_hits %>% 
  filter(aln_length == 101) %>% 
  filter(n_mismatches <= 1) %>%
  filter(n_gap_opens == 0) %>%
  group_by(query_id) %>%
  mutate(n = n()) %>% 
  ungroup() %>% 
  filter(n == 1)
```

```{r}
v6_hits_as_gff <- v6_hits_filtered %>% 
  mutate(chrom = subject_id,
         source = "st_infinium_8303_v4_liftover_to_v6",
         type = "SNP",
         start = s_left + 50,
         end = s_right - 50,
         score = ".",
         strand = "+",
         phase = ".",
         attributes = str_extract(query_id, "solcap_snp_c[0-9]_[0-9]{1,}")) %>% 
  select(chrom,
         source,
         type,
         start,
         end,
         score,
         strand,
         phase,
         attributes)
```

```{r}
nrow(v6_hits_as_gff)
```

> 5,775 SNPs kept after liftover.

```{r}
write_tsv(v6_hits_as_gff, "potato_array_SNPs_v6_liftover.gff3", col_names = F)
```